---
title: "Vignette for the `cir` Package"
subtitle: "Using Up-and-Down data from Benhamou et al. (2003)"
author: "Assaf Oron"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette for the 'cir' Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
#### Note to self: Use this whenever editing the vignette for resubmission/rebuild
#   (after the final CRAN-compatibility check)
# tools::buildVignettes(dir = '.',  tangle=TRUE)
#  (You can also replace the 2 lines below here with point/click directory creation and file copy/paste)
# dir.create("inst/doc")
# file.copy(dir("vignettes", full.names=TRUE), "inst/doc", overwrite=TRUE)
# (from https://community.rstudio.com/t/browsevignettes-mypackage-saying-no-vignettes-found/68656/6 )

options(rmarkdown.html_vignette.check_title = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

## Overview

I demonstrate the `cir` package with data from the anesthesiology experiment of Benhamou et al. (2003)^[Benhamou D, Ghosh C, Mercier FJ. A Randomized Sequential Allocation Study to Determine the Minimum Effective Analgesic Concentration of Levobupivacaine and Ropivacaine in Patients Receiving Epidural Analgesia for Labor. Anesthesiology. 2003;99(6):1383-1386.]. This experiment used the Up-and-Down (UD) dose-finding design, and was re-analyzed a few years later by Pace and Stylianou (2007).^[Pace NL, Stylianou MP. Advances in and Limitations of Up-and-down Methodology: A Précis of Clinical Use, Study Design, and Dose Estimation in Anesthesia Research. *Anesthesiology* 2007;107(1):144-152.] Here I re-analyze the dataset yet again, using state-of-the-art Centered Isotonic Regression (CIR),^[Oron AP, Flournoy N. Centered Isotonic Regression: Point and Interval Estimation for Dose-Response Studies. *Stat Biopharm Res.* 2017;9(3):258-267.] this package's core estimator. 

Up-and-down data analysis was the original motivation for developing CIR and `cir`. An overview of UD designs for general scientific audiences was recently published in Oron et al. (2022).^[Oron AP, Souter MJ, Flournoy N. Understanding Research Methods: Up-and-down Designs for Dose-finding. *Anesthesiology* 2022; 137:137–50. See also the online supplement.] Even more recently, an overview oriented towards statisticians and data scientists was published in Oron and Flournoy (2025).^[Oron AP, Flournoy N. p-and-Down: The Most Popular, Most Reliable, and Most Overlooked Dose-Finding Design, *N Engl J STAT DATA SCI* 3(2025), no. 3, 222-233.] 

Despite the UD connection, the `cir` package is useful for any binary-outcome data for which a monotone dose-response relationship is expected, regardless of how that data were obtained. Given $n$ observation pairs of doses $x_i, i=1, \ldots, n$ and binary responses $y_i, i=1, \ldots, n$, CIR estimates the **dose-response function $F(x)$, or in short, just $F$:** 

$$ F\equiv F(x) \equiv \Pr\left(Y = 1 \mid X = x \right),$$
making no assumptions except for strict monotonicity.


The package includes methods for both **"forward"** estimation of $F$, i.e., the expected response rate conditional on dose, and **"inverse"** estimation of $F^{-1}$, a.k.a. dose-finding. Both estimation directions have quick, user-friendly all-in-one functions, as well as more detailed in-depth functions.

**If your interest is mainly in UD data analysis rather than more general dose-response data, then you are strongly encouraged to use the dedicated UD package `upndown`.** It includes wrappers for even simpler, single-command plotting and estimation of UD datasets, using `cir` package functions with the defaults already dialed onto the UD context. You can download the latest version of that package using `remotes::install_github(assaforon/upndown)`. CRAN also has a copy, but it would always be a couple of steps back, as `upndown` is still new and changing a lot.

## The Study in our Example

Benhamou et al. (2003) randomized pregnant volunteers in labor into two groups, receiving as analgesic either ropivacaine or levobupivacaine, to estimate each agent’s median effective dose (ED$_{50}$) for this condition, and to test whether there is significant difference in equal-concentration efficacy by comparing the two estimates. The original study used a "traditional" UD dose-averaging estimation method, concluding that even though levobupivacaine seemed 19% more potent, the difference between ED$_{50}$s was not significant. To derive inference about this difference, they apparently used the standard errors of dose averages in a $t$-test like manner.

Pace and Stylianou (2007) re-analyzed the data using isotonic regression (IR) and $83\%$ bootstrap confidence intervals (CIs); under certain assumptions, examining whether such CIs overlap is equivalent to rejecting the Null hypothesis with $\alpha = 0.05.$ ^[Payton ME, Greenstone MH, Schenker N. Overlapping confidence intervals or standard error intervals:  What do they mean in terms of statistical significance? *J Insect Sci.* 2003;3(1).] They found a larger difference than Benhamou et al.: according to the IR estimates, levobupivacaine was 37% more potent . Furthermore, the difference was deemed statistically significant because the $83\%$ CIs did not overlap.

Here we revisit this experiment yet again. To reduce some confusion, we drop the percent sign from description of doses used *(in the original, concentrations were described as percents)*.

## Experimental Trajectories

We do not have the original data table, nor do the original authors have access to it anymore; but we can read the sequence of administered doses off of Benhamou et al.’s Figure 1. 


```{r data}
# For brevity, we initially use integers to denote the doses. 

xropi = c(11:9,10:8,9,10,9,10:7,8:11,10:12,11:7,8,7:10,9,8,9,8:10,9,10,9,10)
xlevo = c(11,10,11,10,11:9,10:7,8,7,8:5,6:8,7,8:6,7,6,7,6,7:5,6,7,6:12)
```

Given knowledge of the UD's design rules, the responses ($y$) can be read off directly from the doses ($x$): a positive increment indicates no effectiveness ($y=0$), and vice versa. Symbolically,  $$y = \left( 1-diff(x) \right) / 2.$$ We use this and the `DRtrace()` constructor utility, to create objects that store doses (converted to their physical units) and responses; as we call them here, the experimental "trace" or **trajectory.**

```{r DRtrace}
library(cir)
bhamou03ropi = DRtrace(x = xropi[-40]/100, y = (1-diff(xropi))/2 )
bhamou03levo = DRtrace(x = xlevo[-40]/100, y = (1-diff(xlevo))/2 )
```

In the construction above, we gave up the 40th and last observation in each arm, because we only know its dose but not the response. Since UD datasets (and more generally, small-sample dose-response datasets) are rather compact, as shown above entering the data takes no more than 2-3 lines of code. But you can also use `.csv` file input if preferred, with two columns headed `x` and `y`. 

`DRtrace` objects have a native plotting method:

```{r tracefig,fig.width=9,fig.height=4.5,out.height=400,out.width=800}
par(mfrow=c(1,2), las=1, mar=c(4,4,4,1)) # image format parameters
doserange = c(5,12)/100

plot(bhamou03ropi, ylim=doserange, ylab="Concentration (%)", main='Ropivacaine Arm')
legend('bottomright', legend=c('Effective','Ineffective'), pch=c(19,1), bty='n')
plot(bhamou03levo, ylim=doserange, ylab="Concentration (%)", main='Levobupivacaine Arm')
```

The "traditional" estimates in the original articles, used some type of average of the doses administered. This is based on the premise that over time, up-and-down designs concentrate these doses roughly symmetrically around the target percentile. Generally speaking, the "traditional" estimates are rather outdated and lack robustness. We recommend CIR as the standard estimator for up-and-down experiments (see e.g., [Oron et al. 2022 supplement (pdf link)](https://cdn-links.lww.com/permalink/aln/c/aln_2022_05_25_oron_aln-d-21-01101_sdc1.pdf)).

## Dose-Response Plot and ED$_{50}$ Estimates

To derive CIR estimates, we take the `DRtrace` trajectory objects and generate `doseResponse` dose-rate-count summaries.

```{r doseResponse}
bhamou03ropiRates = doseResponse(bhamou03ropi)
bhamou03levoRates = doseResponse(bhamou03levo)
knitr::kable(bhamou03ropiRates, row.names=FALSE, align='ccr', digits=c(2,4,0) )
knitr::kable(bhamou03levoRates, row.names=FALSE, align='ccr', digits=c(2,4,0) )
```

Let us visualize the response frequencies on **the dose-response plane, which is "where CIR estimation action happens"** *(and likewise for any regression-based estimation of dose-response data)*. 

Analogously to the plots above, the plots below are native `cir` methods for `doseResponse` objects. Symbol area is proportional to the number of observations at each dose. To change to fixed-size, use `varsize=FALSE`.


```{r drfig0,fig.width=10,fig.height=5,out.height=400,out.width=800}
par(mfrow=c(1,2), las=1, mar=c(4,4,4,1)) # image format parameters
plot(bhamou03ropiRates, xlab="Concentration (%)", 
ylab="Proportion Effective", main='Ropivacaine Arm', ylim=0:1)
# Showing the target response rate
abline( h=0.5, col='purple', lty=2)

plot(bhamou03levoRates, xlab="Concentration (%)", 
ylab="Proportion Effective", main='Levobupivacaine Arm', ylim=0:1)

abline( h=0.5, col='purple', lty=2)
```

Target-dose estimation via regression is an [**"Inverse Problem"**](https://en.wikipedia.org/wiki/Inverse_problem): we draw a line at the desired $y$ value (in this case, $50\%$ or 0.5, marked in purple), and look for the best-fitting $x$ value. Indeed, statistically the task is described as estimating $F^{-1}(0.5)$, or more generally $F^{-1}(\Gamma)$, with $100\Gamma$ being the target percentile. 

With the Ropivacaine arm data (left), there is a relatively clear transition between low-response and high-response regions, so the target seems to lie between concentrations $0.09$ and $0.10$. We cannot be very confident of that, given the limited amount of information - but at least there's a clear candidate. 

With Levobupivacaine data (right) the picture is far more murky. Concentrations $0.07$ and $0.08$ go above $50\%$ response, but  $0.09$ and $0.10$ go back below that level. As we visualize further below, CIR helps clear that picture via the simplest pooled response-rate averages that makes the overall dose-response curve obey monotonicity.

But before going behind the scenes: from the `doseResponse` object, CIR estimation of $F(x)$  is only one step away.

```{r fcir}
quickIsotone(bhamou03ropiRates, target=0.5, adaptiveShrink=TRUE)
quickIsotone(bhamou03levoRates, target=0.5, adaptiveShrink=TRUE)
```

Estimating the ED$_{50}$, a.k.a. **the target dose**, is done very similarly:

```{r cir}
ropiTargetCIR = quickInverse(bhamou03ropiRates, target=0.5, adaptiveShrink=TRUE, conf = 0.95)
ropiTargetCIR 
levoTargetCIR = quickInverse(bhamou03levoRates, target=0.5, adaptiveShrink=TRUE, conf = 0.95)
levoTargetCIR
```

You can also do the estimation directly from the `x, y` data; the functions will create a `doseResponse` object on the fly.

The ratio between our point estimates is `r round(ropiTargetCIR$point/levoTargetCIR$point, 2)`, similar to the Pace-Stylianou IR estimate. Not too surprising, as the methods are close to each other than to the original authors' dose-averaging approach, and therefore less affected by starting dose.

### Other Notes
 - The target-dose estimate appears under the word `point`: `r round(ropiTargetCIR$point, 4)` for ropivacaine and `r round(levoTargetCIR$point, 4)` for  levobupivacaine.
 - We specify the target response-rate as a fraction, via the `target` argument.
 - Default CIs are 90% rather than 95%, due to the large variability and limited confidence in these typically small-sample experiments. Here we use the `conf` argument to obtain 95% CIs, which are the more appropriate ones when comparing arms.
 - The `adaptiveShrink` option performs an empirical correction of the bias induced by adaptive designs, a bias described by Flournoy and Oron (2019).^[Flournoy N, Oron AP. Bias induced by adaptive dose-finding designs. J Appl Stat. 2019;47(13-15):2431-2442.] This type of bias is induced not just by UD designs, but also by model-based adaptive designs such as the Continual Reassessment Method. It is minimal near the target dose, but flares out to substantial magnitudes in both directions. **Because of this bias, we strongly recommend to *not* estimate any dose that is far removed from the dose-finding design's designated target, except for exploratory/illustrative purposes.** The correction offered in `cir` serves mostly to provide better CI coverage for the target dose estimate.


## Comparing the Two Arms

Having examined the methodology behind the CI-overlap approach (which to be prudent, should have been between 84% rather than 83% CIs), we do **not** recommend it anymore. Besides its somewhat contrived nature, this method is not robust to deviations from distribution assumptions, or even to sample size. Instead, we recommend the more mainstream approach of calculating the (95%) CI for the difference and observing whether it contains zero. 

To obtain that CI we use the assumption that CI widths are roughly proportional to the square-roots of the variances, as well as independence between arms, while allowing for CI asymmetry, and calculate

$$
\begin{array}{rcl}
LHW_{diff.} & = & \sqrt{LHW_1^2 + RHW_2^2} \\
RHW_{diff.} & = & \sqrt{RHW_1^2 + LHW_2^2}, 
\end{array}
$$
where LHW, RHW denote the left- and right-hand widths respectively, and group 2 is the one whose estimates are being subtracted from group 1. Let us put numbers into these formulae:

```{r ciofdiff}
lhw_1 = ropiTargetCIR$point - ropiTargetCIR$lower95conf
rhw_1 = ropiTargetCIR$upper95conf - ropiTargetCIR$point 
lhw_2 = levoTargetCIR$point - levoTargetCIR$lower95conf
rhw_2 = levoTargetCIR$upper95conf - levoTargetCIR$point 

lhw = sqrt(lhw_1^2 + rhw_2^2)
rhw = sqrt(lhw_2^2 + rhw_1^2)

## Lower 95% bound of difference 
ropiTargetCIR$point - levoTargetCIR$point - lhw
## Lower 95% bound of difference 
ropiTargetCIR$point - levoTargetCIR$point + rhw

```


Somewhat surprisingly, we reach a conclusion similar to the original authors: **the 95% CI for the difference contains zero, suggesting no evidence for a difference in potency** at the $\alpha=0.05$ level. Where we part ways with Pace and Stylianou - dramatically - is indeed the confidence intervals. Not so much the 83% vs. CI-of-difference, but rather the individual CIs themselves for which, per our repeated examinations over the years, the Pace-Stylianou approach lacks coverage. That approach relied upon the bootstarp, whereas we use an analytical approach based on Morris’ theoretical work with intervals for datasets with monotone dose-response data,^[Morris MD. Small-Sample Confidence Limits for Parameters under Inequality Constraints with Application to Quantal Bioassay. Biometrics. 1988;44:1083-1092.]  with modifications.

By the way, we can use the same `quickInverse()` function to reproduce the Pace-Stylianou *point* estimates (but not the CIs, since `cir` does not implement the bootstrap approach):

```{r pacesty}
PSropiEstimate = quickInverse(bhamou03ropiRates, target=0.5, estfun=oldPAVA, conf=0.83)
PSropiEstimate
PSlevoEstimate = quickInverse(bhamou03levoRates, target=0.5, estfun=oldPAVA, conf=0.83)
PSlevoEstimate
```


Note the argument specifying the `oldPAVA()` estimation function (PAVA is the name of the leading algorithm to produce IR estimates; the package default is `cirPAVA()`, i.e., CIR). 

Let us visualize more clearly the CIR estimates, and the rationale supporting wider intervals from this dataset.

## Visualizing CIR on the Dose-Response Plane

To construct the full estimated $F(x)$ curves from IR and CIR, we call `cir`'s *"long-form"* forward-estimation functions: 

```{r forward}
ropiCurveIR = oldPAVA(bhamou03ropiRates, full=TRUE)
ropiCurveCIR = cirPAVA(bhamou03ropiRates, target=0.5, adaptiveShrink=TRUE, full=TRUE)
levoCurveIR = oldPAVA(bhamou03levoRates, full=TRUE)
levoCurveCIR = cirPAVA(bhamou03levoRates, target=0.5, adaptiveShrink=TRUE, full=TRUE)
```

With `full=TRUE`, each of these returns a list of three `doseResponse` objects. Here's one example:
```{r forward2}
levoCurveCIR
```

 - `$output` is the estimates at the original doses *(which is what `quickIsotone()` returns sans the CIs)*, 
 - `$input` is the incoming data, and 
 - `$shrinkage` is the output at the doses where the algorithm makes the pooled estimates (then interpolated to generate `$output`). IOW, **`$shrinkage` is the actual regression curve found by the algorithm.** With `oldPAVA()`, this component will always be identical to `$output`.

We end with the dose-response plots, the regression curves and the target estimates, with the two arms aligned and arranged vertically to help visualize the CI overlap.

**As can be seen below, in `cir` this takes quite a bit of coding. In the new `upndown` package dedicated to up-and-down designs *(currently in beta version on GitHub `assaforon/upndown`, and soon also on CRAN)*, most of what you see in the plot can be achieved in a single command with the `drplot()` utility** (which does of course rely on `cir` functions in the backend).

```{r drfig,fig.width=6,fig.height=12,out.height=1000,out.width=500}
par(mfrow=c(2,1), las=1, mar=c(4,4,4,1)) # image format parameters
plot(bhamou03ropiRates, xlab="Concentration (%)", 
  ylab="Proportion Effective", main='Ropivacaine Arm', 
  ylim=0:1, dosevals = (5:12)/100 )
# Adding IR and CIR lines with the same colors/lines as article’s Fig. 4
lines(y~x, data=ropiCurveIR$output, lty=2)
lines(y~x, data=ropiCurveCIR$shrinkage, col='blue',lwd=2)
# Showing the CIR estimate, and confidence interval as a horizontal line
points(target ~ point, data=ropiTargetCIR, col='purple', pch=19, cex=2)
lines(c(ropiTargetCIR$lower95conf,ropiTargetCIR$upper95conf), rep(0.5,2), col='purple', lwd=2)
# The estimate appearing in Pace and Stylianou 2007, nudged 0.01 units down:
points(I(target-0.01) ~ point, data=PSropiEstimate, cex=2)
lines(c(0.08, 0.1), rep(0.49,2))

# Adding legend:
legend('topleft', legend=c("Observed Proportions", 'Isotonic Regression',
                              'Centered Isotonic Regression', paste(c('CIR', 2007), 'Estimate +/- 95% CI')),
       bty='n',pch=c(4,NA,NA,16,1), col=c('black','black','blue','purple', 'black'), lty=c(0,2,1,1,1))

### Now, second plot for Levobupivacaine
plot(bhamou03levoRates, xlab="Concentration (%)", 
  ylab="Proportion Effective", main='Levobupivacaine Arm', 
  ylim=0:1,  dosevals = (5:12)/100)

lines(y~x, data=levoCurveIR$output, lty=2)
lines(y~x, data=levoCurveCIR$shrinkage, col='blue',lwd=2)
points(target ~ point, data=levoTargetCIR, col='purple', pch=19, cex=2)
lines(c(levoTargetCIR$lower95conf,levoTargetCIR$upper95conf), rep(0.5,2), col='purple', lwd=2)
points(I(target-0.01) ~ point, data=PSlevoEstimate, cex=2)
lines(c(0.058, 0.095), rep(0.49,2))

```

### Notes

 - The 2007 estimates were shifted slightly downward in the plot, to make them visible.
 - Referring back to the "murky" picture around target in the Levobupivacaine dose-response plots, we see that both IR (black dashes) and CIR (solid blue) pool the observations from $x = 0.08, 0,09, 0.10$ into a single weighted average $y$ value, which is only slightly higher than the estimate at $x = 0.07.$ The main difference is that IR creates a flat "stretch" with that pooled $y$ value, whereas CIR also pools along the $x$ axis into a single point on the dose-response plane. So both these nonparametric algorithms *"convert the murkiness"* into concluding (with very limited confidence, given the overall small $n$) that **$F(x)$ is very shallow for a long stretch, right above the $y = 0.5$ target response rate.**
 
 - Prior to `cir` package version 2.3.0, the estimated slope informing the CI was the same to the right and left of target, and hence CIs were usually near-symmetric *(and in this particular example, much shorter)*. We now estimate different *"left"* and *"right"* slopes. To revert to the single-slope version, use the argument `slopeRefinement = FALSE` in your `quickInverse()` call.  
 - In this demonstration the CIR curves differ from the ordinary IR curves not only in the CIR algorithm that “shrinks” horizontal intervals to single points, but also in the bias correction found by Flournoy and Oron (2019) and mentioned earlier. In principle this correction is compatible with both methods, but here we applied it only to CIR, because it did not exist at the time of Pace and Stylianou’s article.


## References
